---
title: "RNA-seq preprocessing and quality control — extended analysis"
author: "Adapted from ENS practical scripts"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
editor_options:
  chunk_output_type: console
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(fig.align="center", fig.width=8, fig.height=6)

suppressPackageStartupMessages({
  library(DESeq2)
  library(dplyr)
  library(reshape2)
  library(ggplot2)
})

theme_set(theme_classic())

# ---------- USER SETTINGS ----------
count_file <- "/shared/projects/tp_2636_usorbonne_geno_transcriptomic_184414/data/rnaseq/count_data_diffAnalysis_no0counts.txt"
design_file <- "/shared/projects/tp_2636_usorbonne_geno_transcriptomic_184414/data/rnaseq/design.txt"

# Keep this FALSE for the first run. Turn TRUE only if you want the optional low-count sensitivity analysis.
run_prefilter_sensitivity <- TRUE

# Example candidate genes; edit freely.
candidate_genes <- c(
  "CPAR2_603610",
  "CPAR2_100710",
  "CPAR2_201490",
  "CPAR2_405010",
  "CPAR2_203450"
)
excluded_gene <- "CPAR2_404850"
```

# 1. Load and validate the data

```{r load}
countData <- read.table(
  count_file,
  row.names = 1,
  header = TRUE,
  check.names = FALSE
)

colData <- read.table(
  design_file,
  row.names = 1,
  header = TRUE,
  check.names = FALSE
)

# Force exact sample matching/order.
stopifnot(all(colnames(countData) %in% rownames(colData)))
colData <- colData[colnames(countData), , drop = FALSE]
stopifnot(identical(colnames(countData), rownames(colData)))

# Make the reference level explicit.
colData$condition <- factor(colData$condition)
if ("N" %in% levels(colData$condition)) {
  colData$condition <- relevel(colData$condition, ref = "N")
}

cat("Count matrix:", nrow(countData), "genes x", ncol(countData), "samples\n")
cat("Samples per condition:\n")
print(table(colData$condition))
cat("Total zeros:", sum(countData == 0), "\n")
```

# 2. Basic library-size QC

```{r library_sizes}
libsize_df <- data.frame(
  sample = colnames(countData),
  library_size = colSums(countData),
  condition = colData$condition
)

ggplot(libsize_df, aes(sample, library_size, fill = condition)) +
  geom_col() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Raw library sizes",
    x = NULL,
    y = "Total mapped counts"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# 3. DESeq2 object and standard normalization

```{r dds}
dds <- DESeqDataSetFromMatrix(
  countData = round(as.matrix(countData)),
  colData = colData,
  design = ~ condition
)

# Canonical DESeq2 pipeline will estimate size factors, dispersions and the Wald model.
dds <- DESeq(dds)

cat("Size factors:\n")
print(sizeFactors(dds))
cat("\nModel coefficients:\n")
print(resultsNames(dds))
```

# 4. Distribution before and after normalization

This version does **not** hard-code `5788` genes or assume a fixed sample order.

```{r boxplots, fig.height=8}
make_long_counts <- function(mat, metadata) {
  x <- melt(log2(mat + 1))
  colnames(x) <- c("ORF", "Samples", "Expr.Val")
  x$Condition <- metadata[as.character(x$Samples), "condition"]
  x
}

Raw.data <- make_long_counts(counts(dds), colData)
Norm.data <- make_long_counts(counts(dds, normalized = TRUE), colData)

ggplot(Raw.data, aes(Samples, Expr.Val, fill = Condition)) +
  geom_boxplot(outlier.size = 0.3) +
  labs(
    title = "Expression distributions before normalization",
    x = NULL,
    y = expression(log[2](count + 1))
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggplot(Norm.data, aes(Samples, Expr.Val, fill = Condition)) +
  geom_boxplot(outlier.size = 0.3) +
  labs(
    title = "Expression distributions after DESeq2 normalization",
    x = NULL,
    y = expression(log[2](normalized~count + 1))
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# 5. Dispersion diagnostics

```{r dispersion}
plotDispEsts(dds)
```

# 6. VST, PCA and sample-level QC

For unbiased exploratory QC, use `blind=TRUE`. We also compute `blind=FALSE` for downstream candidate visualization.

```{r vst}
vsd_blind <- vst(dds, blind = TRUE)
vsd_design <- vst(dds, blind = FALSE)
```

```{r pca, fig.height=7}
pca <- plotPCA(vsd_blind, intgroup = "condition", returnData = TRUE)
percentVar <- round(100 * attr(pca, "percentVar"), 1)

ggplot(pca, aes(PC1, PC2, shape = condition, label = name)) +
  geom_point(size = 4) +
  geom_text(nudge_y = 0.25, check_overlap = TRUE) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  ggtitle("PCA of samples (blind VST)")
```

# 7. Pairwise replicate correlations

```{r correlations, fig.height=8}
vmat <- assay(vsd_blind)

pearson_cor <- cor(vmat, method = "pearson")
spearman_cor <- cor(vmat, method = "spearman")

if (requireNamespace("pheatmap", quietly = TRUE)) {
  pheatmap::pheatmap(
    pearson_cor,
    main = "Pearson correlation between samples"
  )
  pheatmap::pheatmap(
    spearman_cor,
    main = "Spearman correlation between samples"
  )
} else {
  image(pearson_cor, main = "Pearson sample correlation")
  image(spearman_cor, main = "Spearman sample correlation")
}
```

# 8. Sample-to-sample distances

```{r sample_distances, fig.height=8}
sampleDists <- dist(t(vmat))
sampleDistMatrix <- as.matrix(sampleDists)

if (requireNamespace("pheatmap", quietly = TRUE)) {
  pheatmap::pheatmap(
    sampleDistMatrix,
    main = "VST sample-to-sample distances"
  )
} else {
  image(sampleDistMatrix, main = "VST sample distances")
}
```

# 9. Cook's distance diagnostics

High Cook's distance can flag genes strongly influenced by one sample.

```{r cooks}
cooks <- assays(dds)[["cooks"]]
maxCook <- apply(cooks, 1, max, na.rm = TRUE)

cook_df <- data.frame(
  gene = names(maxCook),
  maxCook = maxCook
)

ggplot(cook_df, aes(maxCook)) +
  geom_histogram(bins = 100) +
  scale_x_log10() +
  labs(
    title = "Distribution of maximum Cook's distance per gene",
    x = "Maximum Cook's distance (log10 scale)",
    y = "Genes"
  )

head(cook_df[order(cook_df$maxCook, decreasing = TRUE), ], 20)
```

# 10. Candidate-level normalized expression

```{r candidate_counts, fig.height=5}
candidate_present <- intersect(candidate_genes, rownames(dds))

if (length(candidate_present) == 0) {
  cat("None of the default candidate IDs are present. Edit candidate_genes in the setup chunk.\n")
} else {
  for (g in candidate_present) {
    dat <- plotCounts(
      dds,
      gene = g,
      intgroup = "condition",
      returnData = TRUE
    )
    p <- ggplot(dat, aes(condition, count)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(width = 0.08, height = 0, size = 2.8) +
      scale_y_log10() +
      labs(
        title = paste("Normalized counts:", g),
        x = "Condition",
        y = "Normalized count (log10 scale)"
      )
    print(p)
  }
}
```

# 11. Optional low-count filtering sensitivity

This does not replace the main analysis. It checks whether conclusions are sensitive to weakly informative genes.

```{r filter_sensitivity}
if (run_prefilter_sensitivity) {
  keep_10x3 <- rowSums(countData >= 10) >= 3

  cat("Genes before filtering:", nrow(countData), "\n")
  cat("Genes retained by >=10 counts in >=3 samples:", sum(keep_10x3), "\n")

  countData_f <- countData[keep_10x3, , drop = FALSE]

  dds_f <- DESeqDataSetFromMatrix(
    countData = round(as.matrix(countData_f)),
    colData = colData,
    design = ~ condition
  )
  dds_f <- DESeq(dds_f, quiet = TRUE)

  res_main <- results(dds, contrast = c("condition", "H", "N"))
  res_f <- results(dds_f, contrast = c("condition", "H", "N"))

  common <- intersect(rownames(res_main), rownames(res_f))
  filter_compare <- data.frame(
    gene = common,
    LFC_unfiltered = res_main[common, "log2FoldChange"],
    LFC_filtered = res_f[common, "log2FoldChange"]
  )

  ggplot(filter_compare, aes(LFC_unfiltered, LFC_filtered)) +
    geom_point(alpha = 0.25) +
    geom_abline(slope = 1, intercept = 0, linetype = 2) +
    labs(
      title = "Sensitivity to low-count filtering",
      x = "log2FC: supplied no-zero-count matrix",
      y = "log2FC: >=10 counts in >=3 samples"
    )
}
```

# 12. Reproducibility

```{r reproducibility}
format(Sys.time(), "%d %B, %Y, %H:%M")
sessionInfo()
```
